---
title: "560 Group Project: Descriptive Statistics"
author: "Brittany Jones, Tess Abrahamson-Richards, Natalie Turner, Hanyi Wang"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message = FALSE,warning = FALSE)
library(tidyverse)
library(mlmRev)
library(lme4)
library(arm)
library(haven)
library(table1)
library(Hmisc)
library(ggplot2)
library(ggpubr)
library(expss)
library(dplyr)
library(gridExtra)
library(flextable)

## ------------- working directories for Hanyi ----------
wd_hanyi = '/Users/hanyiwang/Desktop/Social-Service-Participation-Among-Older-Adult-Women-in-Washington-State'
path_hanyi = c("../SIPPdata/sipp.csv")

## ------------- read data ----------
setwd(wd_hanyi)
sipp = read.csv(path_hanyi)

unique = sipp %>% distinct(UQID, .keep_all=TRUE) # unique participants

```

## Population of Interest

Filter by women over 50 years old in the SIPP study, there are `r nrow(sipp)` entries, and `r nrow(unique)` unique participants in our dataset.

## Key Variables

### Outcome: Monthly social service utilization

(The following variables are all binary variables with levels "Yes" or "No". We count each "Yes" as 1 point and calculate the monthly social service utilization for each participant.)

- `RGA_MNYN`: Received GA benefits this month

- `RSNAP_MNYN`: Received SNAP benefits this month

- `RSSI_MNYN`: Received SSI benefits this month

- `RTANF_MNYN`: Received TANF benefits this month

- `RWIC_MNYN`: Received WIC benefits this month

- `ECHLD_MNYN`: Received child care service assistance to go to work or school training this month (1-12)

- `ECVDMNYN`: Received the EIP(s), or stimulus check(s) this month (1-12)

- `EOTHAS_MNYN`: Received other (clothing or housing) assistance this month (1-12)

- `ETRANS_MNYN`: Received transportation assistance this month (1-12)

- `EUC1MNYN`: Received regular, government-provided Unemployment Compensation payments in this month of the reference period (1- 12)

- `EUC2MNYN`: Received supplemental, employer-provided Unemployment Compensation payments in this month of the reference period (1- 12)

- `EUC3MNYN`: Received other Unemployment Compensation payments, including union benefits, in this month of the reference period (1- 12)

```{r}
table1(~ECHLD_MNYN+ECVDMNYN+EUC1MNYN+EUC2MNYN+EUC3MNYN,data = sipp)
```

         
As shown above, variables `ECHLD_MNYN`, `ECVDMNYN`, `EUC1MNYN`, `EUC2MNYN`, `EUC3MNYN` have a lot of missing values, so we exclude them from the calculation of social service utilization scores. Thus the monthly social service utilization score for a participant can be 0 to 7. 

### Covariates fitted in the models

Demographic vairables:

- `ESEX`: sex (all "female" based on our inclusion criteria, so will not be in the models)
                    
- `TRACE`: detailed race

- `EORIGIN`: is Spanish, Hispanic, or Latino, or not

- `EEDUC`: highest level of school completed or the highest degree received by December of (reference year)
               
- `EHLTSTAT`: self-reported health status (Confounder)


Individual's total cost of health care `HI` (sum of the following 2 variables)

- `TMDPAY`: non-premium medical out-of-pocket expenditures on medical care

- `TOTCMDPAY`: non-premium medical out-of-pocket expenditures for over-the-counter health-related products

```{r}
## ------------- modify data set ----------
sipp = sipp %>%
  mutate(SS_SCORE = factor(SS_SCORE)) %>%
  mutate(MONTHCODE = factor(MONTHCODE))
  
## ------------- modify labels and units ----------
unique = apply_labels(unique, 
                    ESEX = "Sex",
                    ERACE = "Race",
                    TRACE = "Detailed race",
                    EORIGIN = "Is Spanish, Hispanic, or Latino",
                    EEDUC = "Highest degree",
                    EHLTSTAT = "Health status")

sipp = apply_labels(sipp, 
                    SS_SCORE = "Monthly social service utilization",
                    HI = "Health care utilization")


units(sipp$HI) = "dollars"
units(sipp$SS_SCORE) = "score"
```

## Descriptive Statistics

```{r}
## ------------- descriptive tables ----------
demographic = table1(~+TRACE+EORIGIN+EEDUC+EHLTSTAT,data=unique, caption = "Demographic charactersitics for older adult women in the SIPP study")

ss_score = table1(~(SS_SCORE),data=sipp, caption = "Social service utilization for older adult women in the SIPP study")

hi = table1(~HI,data=sipp, caption = "Health care spending for older adult women in the SIPP study")

# t1flex(demographic) %>% save_as_docx(path="demographic.docx")

demographic
ss_score
hi
```

```{r out.width="70%",fig.align="center",fig.cap='Social service utilization for different healthcare cost'}
## ------------- SS_SCORE VS HI ----------
ggplot(sipp,aes(x=HI,y=SS_SCORE)) +
  geom_point() +
  labs(y = "Social service utilization score", x = "Healthcare costs")

```

```{r fig.align="center",fig.cap='Histogram of social service utilization for different races'}

## ------------- SS_SCORE VS race ----------
ggplot(sipp, aes(x=(SS_SCORE), color=ERACE,fill=ERACE)) + 
  geom_histogram(position = "dodge", binwidth=2,stat="count") +
  labs(x = "Social service utilization score") +  
  theme(legend.position="top") 

```

We also studied the relationship of social service utilization for each individual and month. We first used a histogram and found that there's no significant difference for the distribution of social service utilization scores over different months. (**Figure 3**)

```{r out.width="70%",fig.align="center",fig.cap='Social service utilization distribution over 12 months'}
## ------------- SS_SCORE VS month ----------
#### boxplot
ggplot(sipp, aes(x=(MONTHCODE), color=SS_SCORE,fill=SS_SCORE)) +
  geom_histogram(position = "dodge", binwidth=2,stat="count") +
  labs(x = "Social service utilization score ") +
  theme(legend.position="top")
```

We then made a plot of all participants' social service utilization scores over months, and found a lot of overlapping. (**Figure 4**)

```{r out.width="70%",fig.align="center",fig.cap='Social service utilization changes over 12 months for all participants'}
#### all individuals
sipp = sipp %>%
  mutate(SS_SCORE = as.numeric(SS_SCORE)) %>%
  mutate(MONTHCODE = as.numeric(MONTHCODE)) %>%
  mutate(UQID = factor(UQID))


ggplot(sipp,  aes(x = MONTHCODE,y = SS_SCORE,color = UQID)) +
  geom_point()+ geom_line() +
  ylim(0,5) + theme(legend.position = "none") + scale_x_continuous(breaks=seq(0,12,by=1)) +
  labs(y = "Social service utilization score",x = "Month")


#### randomly pick individuals
# rd = sample(nrow(unique), 30)
# sipp_rd = sipp[sipp$UQID %in% unique[rd,]$UQID,]
# sipp_rd = sipp_rd %>%
#   mutate(SS_SCORE = as.numeric(SS_SCORE)) %>%
#   mutate(MONTHCODE = as.numeric(MONTHCODE)) %>%
#   mutate(UQID = factor(UQID))
# 
# ggplot(sipp_rd,  aes(x = MONTHCODE,y = SS_SCORE,color = UQID)) +
#   geom_point()+
#   geom_line() +
#   ylim(0,5) +
#   theme(legend.position = "none")

```

```{r out.width="130%",fig.align="center",fig.cap='Social service utilization changes over 12 months for participants with a change'}
#### all individuals in seperate plots

# p = list()
# for (i in 1:30){
#   sipp_rd = sipp[sipp$UQID == factor(unique$UQID)[i+1000],]
#   p[[i]] = ggplot(sipp_rd,  aes(x = MONTHCODE,y = SS_SCORE)) +  
#     geom_point()+ geom_line() + ylim(0,5) + scale_x_continuous(breaks=seq(0,12,by=3))
# }
# 
# do.call(grid.arrange,p)


# after carefully screening the individuals, we found most participants have this trend of having a smaller score around summer, and a larger score over spring and fall. We randomly picked some of the participants shown here as an illustration of the trends.

in_ch = c(9, 6,16,45, 57,319,71,67, 80,94,95,117, 315,316,1013,1015)

p = list()
for (i in 1:length(in_ch)){
  sipp_rd = sipp[sipp$UQID == factor(unique$UQID)[in_ch[i]],]
  p[[i]] = ggplot(sipp_rd,  aes(x = MONTHCODE,y = SS_SCORE)) +  
    geom_point()+ geom_line() + ylim(0,5) + scale_x_continuous(breaks=seq(0,12,by=3)) +
    labs(y = "SS score",x = "Month")

}

do.call(grid.arrange,p)



```


To study the difference in more details, we went ahead to make a single plot for the trend for each participant. After carefully screening, we found most participants have the trend of having a smaller score around summer, and a larger score over spring and fall. We randomly picked some of the participants shown here as an illustration of the trends. (**Figure 5**) This trend further instructed us to put month as a covariate in our models.


## Code Appendix

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE,include=TRUE}
```
