---
title: "560 Group Project: Models"
author: "Brittany Jones, Tess Abrahamson-Richards, Natalie Turner, Hanyi Wang"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message = FALSE,warning = FALSE)
library(tidyverse)
library(mlmRev)
library(lme4)
library(arm)
library(haven)
library(table1)
library(Hmisc)
library(ggplot2)
library(ggpubr)
library(expss)
## ------------- working directories for Hanyi ----------
wd_hanyi = '/Users/hanyiwang/Desktop/Social-Service-Participation-Among-Older-Adult-Women-in-Washington-State'
path_hanyi = c("../SIPPdata/sipp.csv")

## ------------- read data ----------
setwd(wd_hanyi)
sipp = read.csv(path_hanyi)

## ------------- data cleaning ----------
sipp = sipp %>%
  mutate(MONTH.num = as.numeric(MONTHCODE)) %>%
  mutate(UQID = factor(UQID)) %>%
  mutate(HI.log = log(HI + .0001)) # add 0.0001 to avoid 0
```



## Models fitting

- Model 0: Random Intercepts, no covariates.

- Model 1: Random Intercepts, with main IV.

- Model 2: Random Intercepts, with demographic covariates, shared slopes. 

- Model 3: Random Intercepts, with demographic + health status covariates, shared slopes. 

- Model 4: Random Intercepts, with main IV, demographic + health status + time. 

- Model 5: Random Intercepts, with main IV, demographic + health status + time shared + time^2. 

- Model 6: Random Intercepts, with main IV, demographic + health status + time shared + time^2 + varying time. 

- Model 7: Random Intercepts, with main IV, demographic + health status + time shared + time^2 + varying time slopes and time^2 slopes.

```{r}
# Model 0: Random Intercepts, no covariates
M0 <- lmer(SS_SCORE ~ 1 + (1 | UQID), data = sipp)

# Model 1: Random Intercepts, with main IV 
M1 <- lmer(SS_SCORE ~ 1 + (HI.log) + (1 | UQID), data = sipp)

# Model 2: Random Intercepts, with demographic covariates, shared slopes
M2 <- lmer(SS_SCORE ~ 1 + (HI.log) + TAGE + TRACE + EORIGIN + EEDUC + (1 | UQID), data = sipp)

# Model 3: Random Intercepts, with demographic + health status covariates, shared slopes
M3 <- lmer(SS_SCORE ~ 1 + (HI.log) + TAGE + TRACE + EORIGIN + EEDUC + EHLTSTAT + (1 | UQID), data = sipp)

# Model 4: Random Intercepts, with main IV, demographic + health status + time
M4 <- lmer(SS_SCORE ~ 1 + (HI.log) + TAGE + TRACE + EORIGIN + EEDUC + EHLTSTAT + 
             MONTH.num + (1 | UQID), data = sipp)

# Model 5: Random Intercepts, with main IV, demographic + health status + time shared + time^2  
M5 <- lmer(SS_SCORE ~ 1 + (HI.log) + TAGE + TRACE + EORIGIN + EEDUC + EHLTSTAT + 
             MONTH.num + I(MONTH.num^2) + (1 | UQID), data = sipp)

# Model 6: Random Intercepts, with main IV, demographic + health status + time shared + time^2 + varying time slopes  
M6 <- lmer(SS_SCORE ~ 1 + (HI.log) + TAGE + ERACE + EORIGIN + EEDUC + EHLTSTAT + 
             MONTH.num + I(MONTH.num^2) + (1 + MONTH.num | UQID), data = sipp)

# Model 7: Random Intercepts, with main IV, demographic + health status + time shared + time^2 + varying time slopes and time^2 slopes
M7 <- lmer(SS_SCORE ~ 1 + (HI.log) + TAGE + TRACE + EORIGIN + EEDUC + EHLTSTAT +
           MONTH.num + I(MONTH.num^2) + (1 + MONTH.num +I(MONTH.num^2)| UQID), data = sipp)
```

## Models selection based on BIC and AIC

```{r out.width = "70%",fig.align='center',fig.cap='AIC and BIC for the 7 models'}
aic.bic = data.frame(
  models = 0:7,
  df = AIC(M0, M1, M2, M3, M4, M5, M6,M7)$df,
  AIC = AIC(M0, M1, M2, M3, M4, M5, M6,M7)$AIC,
  BIC = BIC(M0, M1, M2, M3, M4, M5, M6,M7)$BIC)
  
knitr::kable(aic.bic,caption = 'AIC and BIC for the 7 models')

ggplot(aic.bic,aes(x = models,y=AIC)) +
  geom_point(aes(x = models,y=AIC,color = "AIC")) +
  geom_line(aes(x = models,y=AIC,color = "AIC")) + 
  geom_point(aes(x = models,y=BIC,color = "BIC")) + 
  geom_line(aes(x = models,y=BIC,color = "BIC")) + 
  labs(x = "models",y = "") +
  scale_colour_manual(name="",values=c("AIC"="#00AFBB","BIC"="#FC4E07")) +
  scale_x_continuous(breaks=seq(0,7,by=1))

```

## Results 

### Interpretion

```{r echo = TRUE}
# # point estimates of shared coefficients
# fixef(M6)

## Within-group variance (residual variance)
sigma(M6)^2

# between-group variance: 0.4957805^2
VarCorr(M6)

## Between-school variance
# variance of sample means

```


### p-values 

```{r fig.align='center',fig.cap='Model parameters and CIs',out.width="70%"}
# summary(M6) # find t values
pvalues = data.frame( 
  covariates = c("log(Health costs)","Month","Month^2",
                 "Race: Black alone","Race: Residual","Race: White alone"),
  coefficient = c(-4.545e-03,3.641e-02,-2.855e-03,1.258e-01,6.292e-03,-7.149e-02 ),
  pvalues = c(2*pt(-8.232,df=39-1) ,2*pt(34.921,df=39-1), 2*pt(-36.579,df=39-1),
              2*pt(5.224,df=39-1) ,2*pt(0.189,df=39-1) ,2*pt(-3.451,df=39-1)))

pvalues$pvalues = formatC(pvalues$pvalues, format = "e", digits = 3)
knitr::kable(pvalues,caption = 'p values for covariates of interest')

# fixef(M6)

ptEstAndCIs <- data.frame(
  coef(M6)$UQID[,"(Intercept)"],
  coef(M6)$UQID[,"(Intercept)"]-se.ranef(M6)$UQID[,"(Intercept)"]*1.96,
  coef(M6)$UQID[,"(Intercept)"]+se.ranef(M6)$UQID[,"(Intercept)"]*1.96
)

colnames(ptEstAndCIs) <- c("RE_ests","RE_L","RE_U")

plot(x = 1:nrow(ptEstAndCIs),
     y = ptEstAndCIs$RE_ests,
     ylim = range(ptEstAndCIs$RE_L, ptEstAndCIs$RE_U),
     xlab = "Participants",
     ylab = expression(paste(alpha[j])),
     pch = 20)
segments(x0 = 1:nrow(ptEstAndCIs),
         y0 = ptEstAndCIs$RE_L, y1 = ptEstAndCIs$RE_U, lwd = 2)

```


### SS_SCORE over months

the density of social score utilization minimum points for all participants

```{r fig.align='center',fig.cap='Density of social score utilization minimum points for individuals'}
axis.sym.vec = -(coef(M6)$UQID[27])/(2*coef(M6)$UQID[28])
axis.sym.vec = data.frame(peak = array(axis.sym.vec[[1]]))

hist(axis.sym.vec$peak,xlim=c(0,12),xlab = "",main = "")
plot(density(axis.sym.vec$peak),xlim=c(0,12),xlab = "Month",main = "")

```

Mean of social utilization scores for each month calculated from the dataset, compared to the prediction of each month’s mean SS scores for a (orange) white woman who’s not spanish, hispanic or latino, with a highest degree of 7th or 8th grade, in good health status, and has a health cost of 1230 (mean value) and an age of 67 (mean value)


```{r fig.align='center',fig.cap="Mean of SS scores for each month based on the data and model prediction"}
months.mean = c()
for (i in 1:12){months.mean[i] = sipp %>% filter(MONTHCODE==i) %>% dplyr::select(SS_SCORE) %>% mean() }
 
# sipp %>% dplyr::select(HI) %>% mean()
# sipp %>% dplyr::select(TAGE) %>% mean()
# summary(M6)[["coefficients"]]
month = c(1,2,3,4,5,6,7,8,9,10,11,12)
model.pred = 0.258358491 - 0.004544700 * (log(1230)) - 0.002060997*(67) -0.071489698 - 0.050425209 + 0.253066929 + 0.172647803 - 0.036408540* month + 0.002855227*(month^2)

ss.mean = data.frame(
  months = 1:12,
  data = months.mean,
  model = model.pred)

ggplot(ss.mean,aes(x = months,y=data)) +
  geom_point(aes(x = months,y=data,color = "data")) +
  geom_line(aes(x = months,y=data,color = "data")) + 
  geom_point(aes(x = months,y=model,color = "model prediction")) + 
  geom_line(aes(x = months,y=model,color = "model prediction")) + 
  labs(x = "Months",y = "Mean SS Scores") +
  scale_colour_manual(name="",values=c("data"="#00AFBB","model prediction"="#FC4E07")) + 
  xlim(0,12)+scale_x_continuous(breaks=seq(0,12,by=1))+ 
  ylim(0,6)+ theme(legend.position = "top")

# plot(months.mean,type = "o",xlim=c(0,12),xlab = "Month",ylab = "SS Score",main = "")

```



\newpage
## Code Appendix

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE,include=TRUE}
```
